---
title: "Reproducible Research: Peer Assessment 1"
author: "Vladislav Mozgovoy"
date: "7/31/2020"
output: 
  html_document:
    keep_md: true
---


---

### Load libraries
```{r, echo=TRUE, results='hide', warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
library(scales)   
```


## Loading and preprocessing the data
Download and unzip file

```{r, echo=TRUE}
if (!file.exists(paste0(getwd(), '/data/activity.csv')) ){
  fileUrl <- "https://d396qusza40orc.clouStepsPerDayront.net/repdata%2Fdata%2Factivity.zip"
  download.file(fileUrl, destfile = paste0(getwd(), '/repdata%2Fdata%2Factivity.zip'))
  unzip("repdata%2Fdata%2Factivity.zip",exdir = "data")
}
StepsPerDay <- read.csv(paste0(getwd(), '/data/activity.csv'), stringsAsFactors = FALSE)
StepsPerDay$date <- as.Date(StepsPerDay$date)
str(StepsPerDay)
```



## Make histogram of the total number of steps taken each day

```{r, echo=TRUE}
stepsByDay <- StepsPerDay %>% group_by(date) %>% summarize(steps = sum(steps, na.rm=TRUE))
head(stepsByDay)
hist(stepsByDay$steps, main="Steps by Day", xlab= "Steps", ylab="Number of Days", breaks=16, col="#E4E4E4")
```

Note that we ignored NA values with na.rm=TRUE

## What is mean total number of steps taken per day?

The mean number of steps per day is
```{r, echo=TRUE}
mean(stepsByDay$steps)
``` 

Add these value to the histogram 

```{r, echo=TRUE}
stepsByDay <- StepsPerDay %>% group_by(date) %>% summarize(steps = sum(steps, na.rm=TRUE))
hist(stepsByDay$steps, 
     main="Steps by Day", 
     xlab= "Steps", 
     ylab="Number of Days",
     breaks=16, 
     col="#E4E4E4")
abline(v=mean(stepsByDay$steps), lty=3, col="#FF4E32")
text(mean(stepsByDay$steps),6,labels="mean", pos=1, col="#FF3232") 
```



## What is the average daily activity pattern?

Summarize data by interval

```{r, echo=TRUE}
stepsByInterval <- StepsPerDay %>% group_by(interval) %>% summarize(steps = mean(steps, na.rm=TRUE))
head(stepsByInterval)
```

And visualize it in plot

```{r, echo=TRUE}
ggplot(data=stepsByInterval, aes(x=interval, y=steps)) +
  geom_line() +
  xlab("5min Interval") +
  ylab("Steps") + 
  scale_y_continuous(labels=comma) + 
  ggtitle("Average Daily Activity Pattern")
```



## Imputing missing values

#### Calculate number of missing values

Total number of rows with 1 or more NA
```{r, echo=TRUE}
sum(!complete.cases(StepsPerDay))
```

We will make a new variable 'StepsNew' that will fill in the missing values of 'steps' with the overall average for the interval

At first we will summarize the data by interval

```{r, echo=TRUE}
 stepsByInterval <- StepsPerDay %>% group_by(interval) %>% summarize(avgsteps = mean(steps, na.rm=TRUE)) %>% ungroup()
 head(stepsByInterval)
```

Then merge it back to original dataframe

```{r, echo=TRUE} 
StepsPerDay <- merge(StepsPerDay, stepsByInterval, by="interval", all.x = TRUE)
head(StepsPerDay)
```

Then calculate 'StepsNew'

```{r, echo=TRUE} 
 StepsPerDay <- StepsPerDay %>% mutate(StepsNew = ifelse(is.na(steps), avgsteps, steps))  
head(StepsPerDay)
```

And at the end we will create new dataset which also includes the new data

```{r, echo=TRUE} 
StepsPerDay2 <- StepsPerDay %>% select(interval, date, StepsNew) %>% rename(steps = StepsNew)
head(StepsPerDay2)
```

Now we can make histogram of the total number of steps taken each day where missing values included

```{r, echo=TRUE} 
 StepsNewByDay <- StepsPerDay %>% group_by(date) %>% summarize(StepsNew = sum(StepsNew, na.rm=TRUE))
 hist(StepsNewByDay$StepsNew,  breaks=16, col="#E4E4E4", main="New Histogram of Steps by Day", xlab= "Steps By Day", ylab="Number of Days")
```

## Are there differences in activity patterns between weekdays and weekends?

Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

```{r, echo=TRUE}
StepsPerDay <- StepsPerDay %>% mutate(day_type = ifelse(weekdays(StepsPerDay$date) %in% c("Saturday", "Sunday"), "Weekend", "Weekday"))
head(StepsPerDay)
```

Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis)

```{r, echo=TRUE}
stepsByInterval <- StepsPerDay %>% group_by(day_type, interval) %>% summarize(steps = mean(steps, na.rm=TRUE)) %>% ungroup()
ggplot(data=stepsByInterval, aes(x=interval, y=steps)) +
  geom_line() +
  xlab("5-Minute Interval") +
  ylab("Steps") + 
  scale_y_continuous(labels=comma) + 
  ggtitle("Average Steps") + facet_grid(day_type ~ .)
```
